The algebraic hierarchy in type theory using type classes

### 2010-06-04 jar.tex
Snapshot of introduction/preliminaries/bundling sections rewrite (*heavy* WIP).

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi\ $}}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
}

@@

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation.

@@

\section{The Type-Classified Algebraic Hierarchy}\label{classes}
Unlike Haskell's and Isabelle's second class type classes, Coq's type classes are first class~\cite{DBLP:conf/tphol/SozeauO08}: classes and their instances are realized as ordinary record types (``dictionaries'') and registered constants of these types.

We represent each structure in the algebraic hierarchy as a type class. This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields. By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e. carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields. Type classes defined in this way are essentially predicates with automatically resolved proofs.

Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records). These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}. Section~\ref{univ} abstracts to universal
algebra, we develop the basic theory to the point of the first homomorphism theorem.
Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
this function is implemented in Coq's tactics language. A more convenient way is to use type class
unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
between type classes and canonical structures.

@@

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}.

The Coq type theory lacks quotient
types, as it would make type checking undecidable. Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called \lstinline|Proper|.
As we will see in Section~\ref{univ} below this functionality can also be used to work with
quotient algebras.

Coq has two kinds of sorts \lstinline|Prop| for propositions and \lstinline|Type| for data types. Coq
is scheduled to support proof irrelevance, every \lstinline|p,q:P:Prop| are
convertible, in the future; see section~\ref{PI}.

@@

\section{To bundle or not to bundle? (Or: What to bundle?)}\label{bundle}
We provide a rationale for our approach. We begin gently.

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we might define:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}

and similar for transitive and symmetric relations.

With this approach, we run into problems when defining \lstinline|EquivalenceRelation| in terms of
\lstinline|ReflexiveRelation| and its two siblings, because when posit three objects of type
\lstinline|ReflexiveRelation|, \lstinline|TransitiveRelation|, and \lstinline|SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation. These equalities
cause difficulties, because the three carriers and relations won't be convertible.
Technically, this leads to the research in manifest fields, i.e.\ the \emph{with} construction, to
which we return in Section~\ref{manifest}.

As a first step we would define \lstinline|ReflexiveRelation| as:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}

This would solve the problem for the carrier, but \lstinline|EquivalenceRelation| would still need equalities for the relation:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_refl: ReflexiveRelation er_carrier
    ; er_sym: SymmetricRelation er_carrier
    ; er_trans: TransitiveRelation er_trans
    ; same_rel: rr_rel er_refl = sr_rel er_sym
    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
\end{lstlisting}

A natural solution is to pull the relation out of \lstinline|ReflexiveRelation| (and its siblings) as well:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{lstlisting}

or equivalently:
\begin{lstlisting}
  Definition ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
    Π x, rr_rel x x.
\end{lstlisting}

This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from \lstinline|Type| to \lstinline|Prop| reflects the fact that \lstinline|ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opaque definitions. The former can be unfolded,
whereas this is impossible for the latter. Since we never need to distinguish between two proofs of
a propositions, propositions can be opaque.}
we will return to this issue in Section~\ref{PI}.
We can now define \lstinline|EquivalenceRelation| without awkward equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}

But looking at \lstinline|EquivalenceRelation|, we see that it now essentially has the same form as the
original naive \lstinline|ReflexiveRelation| record we started with. Clearly, equivalence relations
and reflexive relations are analogous concepts, and by the same argument we
conclude that \lstinline|EquivalenceRelation| should really be defined as:

\begin{lstlisting}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: relation er_carrier): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}

After all, when we define a type with
two different equivalence relations on it,
such as the quotient of a setoid,
we have the same considerations as for \lstinline|ReflexiveRelation| above.

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of \lstinline|Monoid| that bundles the
relation. Then if we proceed to define \lstinline|SemiRing| as a double \lstinline|Monoid|, we face a dilemma:
\begin{lstlisting}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, H
\end{lstlisting}

Do we use \lstinline|mult_monoid|'s equality or \lstinline|plus_monoid|'s equality in \lstinline|H|?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

@@

Another benefit of our approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, to express that some collection of relations and
operations possess the structure as a type class constraint seems to require manifest fields; see
Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

Leaving the equivalence proofs bundled means that when one has a proof that a structure
is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
equivalence. Fortunately, this redundancy is entirely harmless since our terms
never refer (directly or indirectly) to proofs, which is the case when the relations and operations
they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
irrelevance.

From the discussion so far, we derive some principles:
\begin{enumerate}
 \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
(i.e. carriers, relations, and operations) as parameters, and nothing else;
 \item (as a consequence) these concepts should be defined as predicates (which includes records and
classes living in Prop), and their proofs may be kept opaque;
 \item (as a second consequence) algebraic expressions need/should never refer (directly or
indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
famously does refer to a proof that the element is non-zero, but not to a proof that the operations
indeed form a field.}
\end{enumerate}
% \setcounter{enumi}{3}

The principles listed so far are not particularly controversial, in fact this
approach ensures maximum flexibility~\cite{Hints}. However,
there is a common perception that unbundling carriers, relations, and operations this way invariably
leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
problems establishing canonical names for operations (since we no longer project them out of
structure instances all the time).

In the next sections, we show how systematic use of type classes (and their support infrastructure
in Coq), combined with the use of a special form of type class we call an `operational type class'
(as opposed to a `structural' type class we might use in place of the records shown thus far), lets
us avoid these problems.

>>>>

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi\ $}}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot\ $}}1
  { ≡ }{{$\equiv\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
}

@@

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g. rewriting).

@@

% ideally should be more or less readable by Haskell person. that would be neat. so that's what the preliminaries should prepare for.

\paragraph{Outline}
In section \ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section \ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, as this is arguably the biggest design dimension when building interfaces for abstract structures. Having derived some principles concerning bundling, we then proceed to show (in section \ref{managing}) how type classes may be employed to adopt these principles without compromising ease of use. Next, in section \ref{hierarchy}, we introduce a number of additional conventions that let us complete a consistent idiom for interfaces to algebraic structures. With this in place, we continue with a discussion of

to the

% outline:
% 
% - preliminaries
%     -> coq (couple of sentences about programming in a combined framework with logic)
%     -> type classes
% - on bundling
%     -> introduce problem
%     -> naive solution
%     -> etc
%     -> "we first begin by showing how unbundling is in some sense the "right" way to do things"
%     in doing this, however, we will find that doing things this way is somewhat painful
%     -> derived principles
% - Managing unbundled structure with type classes
%     -> painful if flexible, which is why not done before
%     -> with type classes, managable.
% 
% - toward the algebraic hierarchy
% 
%   leaving things unbundled is one thing, but we need more for the alg. hierarchy. suppose we define a class for semigroups based on yada.
%   -> operational type classes (parameterized over a single operation = constant), structural type classes (parameterized over all the components that make up the structure).
% 
% - advantages
% 
% - numeric classes
% 
% - cat.theory
% 
% - univ. alg.
% 
% - quoting

%todo: have a word somewhere about "super classes"

@@

\section{Preliminaries}
\label{preliminaries}

% ideally, this should make the paper readable for Haskell people

The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}. In true Curry-Howard fashion, it can be thought of either as an excessively pure (if somewhat pedantic) functional programming language with an extremely expressive type system, or as a language for mathematical statements and proofs. We highlight some aspect of Coq that are of particular relevance to our development.

\paragraph{Types and propositions}

Propositions in Coq are types, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort of types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (i.e. proof terms).

Adopting this regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations. On a very practical level, this lets the extraction mechanism safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.

A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.

\paragraph{Equality, setoids, and rewriting}

The ``native'' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where ``\lstinline|a ≡ b|'' is notation for \lstinline|eq T a b|~\footnote{We diverge from Coq traditional and reserve a single equals sign for setoid equality.}. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object (for instance when identifying formal fractions of integers representing the same ratio). Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

Effectively keeping track of, resolving, and combining proofs of equivalence-ness and propriety when the user attempts to substitute a given (sub)term using a given equality, requires nontrivial infrastructure and support from the system. The Coq implementation of these mechanisms was largely rewritten in 200? by Matthieu Sozeau \ref{rewriting}. (todo: say that it's now much better than before)

\paragraph{Type classes}

Type classes have been a great success story in the Haskell functional programming language, as a means for organizing interfaces of abstract structures. The type class implementation that is now part of Coq provides a superset of their functionality, but implemented in a different way.

In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus. By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (``dictionaries''), instances are ordinary constants of these record types (registered as ``hints'' with the proof search machinery), class constraints are (fairly) ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.

Thus, type classes in Coq are realized by relatively minor syntactic aides that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.

% Often, we will want to use a \emph{different} equivalence relation when representing algebraic structures, for instance when representing integers by the hypothetical subtraction of a pair of natural numbers. A type and an equivalence relation over that type together form a \emph{setoid}.
% 
% An extremely important
% 
% 
% when doing algebra, however, one wants to use custom equivalence relations, for instance when
% 
% , for the purposes of doing algebra.
% 
% where term reduction is
% 
% the native
% 
% Leibniz equality 
% 
% (which in the Coq system amounts to
% 
% the ``native'' equality is term equality modulo reduction.
% 
% which amounts to 
% 
% normal forms
% 
% A trivial example being the method of representing integers by hypothetical subtractions of one natural from another.
% 
% 
% The Coq type theory lacks quotient
% types, as it would make type checking undecidable. % todo: reference
% Instead it supports setoids (aka Bishop sets), a pair of a type and an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support for setoids has been added to the system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called \lstinline|Proper|.
% As we will see in Section~\ref{univ} below this functionality can also be used to work with
% quotient algebras.

% \section{The Type-Classified Algebraic Hierarchy}\label{classes}
% We represent each structure in the algebraic hierarchy as a type class. This immediately leads to the familiar question of which components of the structure should become parameters of the class, and which should become fields. By far the most important design choice in our development is the decision to turn all \emph{structural} components (i.e. carriers, relations, and operations) into parameters, keeping only \emph{properties} as fields. Type classes defined in this way are essentially predicates with automatically resolved proofs.
% 
% Conventional wisdom warns that while this approach is theoretically very flexible, one risks extreme inconvenience both in having to declare and pass around all these structural components all the time, as well as in losing notations (because we no longer project named operations out of records). These legitimate concerns will be addressed in Sections~\ref{bundle}-\ref{classes}. Section~\ref{univ} abstracts to universal
% algebra, we develop the basic theory to the point of the first homomorphism theorem.
% Section~\ref{interfaces} provides interfaces to usual data structures by providing their universal
% properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
% this function is implemented in Coq's tactics language. A more convenient way is to use type class
% unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparison
% between type classes and canonical structures.

@@

\section{Bundling is bad}\label{bundling}

Algebraic structures are expressed in terms of a number of carriers, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy. In a system like Coq, we have different options when it comes to representing the grouping of these components. On one end of the spectrum, we can simply define the (conjunction of) laws as a $n$-ary predicate over $n$ components, basically forgoing grouping altogether. For instance, for the mundane example of a reflexive relation, we could use:

\begin{lstlisting}
Definition reflexive {A} (R: relation A): Prop := Π a, R a a.
\end{lstlisting}
(The curly brackets used for \lstinline|A| mark it as an implicit argument.)

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in theory (todo: ref), in practice \emph{naive} adoption of this approach in Coq (i.e. without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or resort to awkward existentials. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate convenient notations with them.

% hm, i used to have paragraphs here about how it would be unclear how this approach could support structure inference and structure inheritance, but those are actually pretty straightforward using ordinary proof search.

% hm, say something about opacity

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today (namely CoRN\ref{corn} and Ssreflect\ref{ssreflect}), both use \emph{bundled} representations: records with one or more of the components as fields. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
Record ReflexiveRelation: Type :=
  { rr_carrier: Type
  ; rr_rel: relation rr_carrier
  ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it\footnote{While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``+'' and ``*'' notations to the record field projections for the addition and multiplication operations, respectively.}.

% While not observable in this example, when combined with other specialized facilities in Coq (namely coercions and/or canonical structures), this bundled representation also lets one address the problems of structure inference inheritance to some extent.

Unfortunately, despite its virtues, the bundled representation introduces serious problems of its own, the most immediate and prominent one being a lack of support for \emph{sharing} components between structures.

In our example, lack of sharing support rears its head as soon as we try to define \lstinline|EquivalenceRelation| in terms of \lstinline|ReflexiveRelation| and its hypothetical siblings bundling symmetric and transitive relations. There, we would need some way to make sure that when we bundle the three constituents of types \lstinline|ReflexiveRelation|, \lstinline|SymmetricRelation|, and \lstinline|TransitiveRelation|, respectively together in a record, they all refer to the same carrier and relation. Positing additional equalities between the three bundled carriers and relations is neither easily accomplished (because one would need to work with heterogenous equality) nor would it permit natural use of the resulting structure (because one would constantly have to rewrite things back and forth).

Manifest records \ref{manifestrecords} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by employing type classes instead of bundled records.

If we revert back to the predicate formulation of reflexive (et.al) relations, we \emph{could} define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
\begin{lstlisting}
Record EquivalenceRelation: Type :=
  { er_carrier: Type
  ; er_rel: relation er_carrier
  ; er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}
However, the exact same argument inevitably leads to the conclusion that \lstinline|EquivalenceRelation|, too, should be a predicate. Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.

Another attempt to regain some grouping might be to bundle the carrier with the relation into a record. Unfortunately, this leads to the same problem of sharing, in different clothes: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles. To give a slightly less silly example of this phenomenon: bundling a monoid's constant and binary operation into a record means that when defining a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant, one needs awkard hacks to establish equality of the two operation bundles' constants.

A second problem with bundling components is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (although admittedly coercions and notations can make this less painful).

% hm, what to say about terms referring on proofs.

Thus, bundled representations come at a substantial cost in flexibility. Historically, this cost has been acceptable because (1) the unbundled alternative was such a pain, and (2) the standard algebraic hierarchy (up to, say, fields and modules) is not all that wild.

In the next section, we show that type-classification of structure predicates and their component parameters have the potential to remedy the problems traditionally associated with their use, giving us the best of both worlds.

For completeness, we mention an alternative record-based approach known as ``packed classes'', as used in the ssreflect library. In the packed classes idiom, each algebraic structure is represented by not one but a trinity of interdependent records along with packing/unpacking functions. These are then stacked in an intricate manner designed to reach a particular hybrid packing balance that, combined with the use of canonical structures and coercions, accomplishes some of the goals above.

We have found it hard to assess the general merits of this approach, partly because of its complexity, and partly because the only actual usage example (the hierarchy in the ssreflect library) is not only wedded to a very specialized view on equality and decidability that does not support setoid relations from the outset, but also diverges significantly from standard mathematical dichotomy, making it hard to recognize familiar structures and presentation. The reasons for this (as we understand them) are pragmatic and twofold.

First, the ssreflect library was originally developed strictly to support very practical work in a specialized domain of mathematics involving decidable and mostly finite structures. Second, at the time of ssreflect's initial development, support for setoid rewriting in Coq was not nearly as stable and refined as it is today, which made ``Leibniz equality at all costs'' a more reasonable strategy at the time than it might be considered today.

Work to extend the ssreflect framework with techniques to allow representation of a less restricted class of mathematical structures has only recently begun and is still ongoing, though there are currently no plans to support undecidable structures.

Because in our development we will be working with computable reals, equality on which can never be either Leibniz or decidable, we conclude that the ssreflect library is not usable for us.

% todo: mention proof irrelevance here somewhere, and terms that depend on proofs.

@@

% Another benefit of our approach is that it works very smoothly when these records are in fact type
% classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
% structure record/class, to express that some collection of relations and
% operations possess the structure as a type class constraint seems to require manifest fields; see
% Section~\ref{manifest}. With
% our unbundled approach, on the other hand, such type class constraints are easily expressed without
% any extensions as type inhabitation queries
% % (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
% of types living in Prop, so that we don't even care what instances are found (thus embracing proof
% irrelevance).

% Leaving the equivalence proofs bundled means that when one has a proof that a structure
% is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
% equivalence. Fortunately, this redundancy is entirely harmless since our terms
% never refer (directly or indirectly) to proofs, which is the case when the relations and operations
% they are composed from are left unbundled. Here, too, we enjoy the warm embrace of proof
% irrelevance.
% 
% From the discussion so far, we derive some principles:
% \begin{enumerate}
%  \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
% (i.e. carriers, relations, and operations) as parameters, and nothing else;
%  \item (as a consequence) these concepts should be defined as predicates (which includes records and
% classes living in Prop), and their proofs may be kept opaque;
%  \item (as a second consequence) algebraic expressions need/should never refer (directly or
% indirectly) to proofs about the algebraic structure.\footnote{The reciprocal in a field
% famously does refer to a proof that the element is non-zero, but not to a proof that the operations
% indeed form a field.}
% \end{enumerate}
% % \setcounter{enumi}{3}
% 
% The principles listed so far are not particularly controversial, in fact this
% approach ensures maximum flexibility~\cite{Hints}. However,
% there is a common perception that unbundling carriers, relations, and operations this way invariably
% leads either to unmanageably big algebraic expressions (involving endless projections and/or implicit
% arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
% problems establishing canonical names for operations (since we no longer project them out of
% structure instances all the time).
% 
% In the next sections, we show how systematic use of type classes (and their support infrastructure
% in Coq), combined with the use of a special form of type class we call an `operational type class'
% (as opposed to a `structural' type class we might use in place of the records shown thus far), lets
% us avoid these problems.
